---
title: "Climate Changes Affect on Plant Species"
author: "Lukas Robinson"
date: "3/11/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Climate change here on Earth is at an alarming stage, with a steady increase in atmospheric carbon dioxide affecting how plants photosynthesis to the unbalanced situations in climatic conditions, some places prolonged droughts and in others increased flooding is causing major problems for plant species. Climate change is defined as a long-term change in the average weather patterns that have come to define Earth’s local, regional and global climates. Just with this definition, we can hypothesis that as climate change slowly changes Earth’s local, regional and global climates plant species will be affected in some way. In this research project, I will be answering how climate change is affecting these plant species based on their plant distribution over time. To define plant distribution I will be looking at three different time periods and three common plant species found in Arizona comparing the abundance and distribution of those plant species in that area. I will be using the Botanical Information and Ecology Network (BIEN) package to look at all of the data and the BIEN data will be the data presented.

## loading packages & data set
```{r}
library(tidyverse)
library(BIEN)
library(ape)
library(maps)
library(sp)
```




