---
title: "Climate Changes Affect on Plant Species"
author: "Lukas Robinson"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = FALSE, message=FALSE, warning=FALSE)
```

## R Markdown

## Abstract
Climate change here on Earth is at an alarming stage, with a steady increase in atmospheric carbon dioxide affecting how plants photosynthesis, to the unbalanced situations in climatic conditions, some places prolonged droughts and in others increased flooding. These issues are causing major problems for plant species worldwide. Climate change is defined as a long-term change in the average weather patterns that have come to define Earth’s local, regional and global climates. Just with this definition, we can hypothesis that as climate change slowly changes Earth’s local, regional and global climates plant species will be affected in some way. In this research project, I will be answering how climate change is affecting these plant species based on their plant distribution over time. To define plant distribution I will be looking at three different time periods and three common plant species found at different elevations in Arizona and California then comparing the abundance and distribution of those plant species in that area. I will be using the Botanical Information and Ecology Network (BIEN) package to look at all of the data and the BIEN data will be the data presented. 

## Introduction

## Methods

### Data Acquisition

For this project, all the data was analyzed and manipulated using RStudio (RStudio Team 2020). I obtained data from the Botanical Information and Ecology Network (BIEN, 2021). This data was acquired using the R package known as BIEN (Maitner, 2020). BIEN was used to obtain information about plant distribution and abundance over numerous years. No special access was required to access this resource I just needed to download the package into R Studio before looking at the data. Since the data from the BIEN package is so large I narrowed down my search for a good research project by searching for three plant species using information found from azstateparks.com. 

### Data Preparation

After first loading in all of the required packages, my data was first prepared for analysis by filtering through all the plant species in the BIEN package to only the three plant species that I wanted to look at. 

Once I filtered out my three species I then bound these three data sets together into one and simplified that data set to only have the categories I needed for my research. The categories include species, date collected, year, and geometry. By doing this it helped me better understand how large my sample size will be for this experiment.

After I created this new data set, I filtered out the two states that I will be looking at, California and Arizona. Once that was done I graphed the data into three separate rows and columns. This showed what the distribution looked like of the three plant species over the three given years.

Next, I used the elevatr package to help me get each elevation point for the plant species found in the data set. This was done by filtering each plant species individually then binding those data sets together. Then calculating the mean, standard deviation, sample size, standard error, and the upper and lower limit. This allowed me to then be able to plot these plant species and see what each of their mean and their 95% confidence interval look like over time and how it has changed. 

The same process was done to get the latitude graph over time. The only difference was plugging in latitude instead of elevation. I then graphed this to see how the mean latitude of these species has changed.

## Results

```{r}
library(tidyverse)
library(BIEN)
library(ape)
library(maps)
library(sp)
library(lubridate)
library(sf)
library(elevatr)
library(USAboundaries)
```

```{r}
Encelia_farinosa <-
  BIEN_occurrence_species(species = "Encelia farinosa")

Rafinesquia_neomexicana <-
  BIEN_occurrence_species(species = "Rafinesquia neomexicana")

Phacelia_distans <-
  BIEN_occurrence_species(species = "Phacelia distans")
```

```{r}
all_data <-
  bind_rows(Encelia_farinosa,
            Rafinesquia_neomexicana,
            Phacelia_distans) %>% 
  as_tibble() %>% 
  transmute(
    species = scrubbed_species_binomial,
    latitude,
    longitude,
    date_collected,
    year = as.factor(year(date_collected))
  ) %>% 
  filter(year %in% c(2015, 2005, 1995)) %>% 
  st_as_sf(coords = c("longitude", "latitude"),crs = 4326, remove = FALSE) %>% 
  print()
```

```{r}
map_data("state") %>% 
  filter(region == "arizona") %>% 
  st_as_sf(coords = c("long", "lat"),crs = 4326, remove = FALSE)
```

```{r}
us_states() %>% 
  filter(state_name %in% c("Arizona", "California")) %>% 
  ggplot() + 
  geom_sf() +
  geom_sf(mapping = aes(color = species, shape = year),
          data = all_data) +
  facet_grid(year ~ species) +
  guides(color = "none") +
  guides(shape = "none")
```

Figure 1. A map displaying the distributions of the three plant species in Arizona and California for the years 1995, 2005, 2015. 

```{r}
elevation_point<-all_data %>% 
  get_elev_point() 
```

```{r}
Encelia_elevation <-
  filter(elevation_point, species %in% c("Encelia farinosa")) 

Phacelia_elevation <-
  filter(elevation_point, species %in% c("Phacelia distans"))

Rafinesquia_elevation <-
  filter(elevation_point, species %in% c("Rafinesquia neomexicana"))
```

```{r}
all_elevation <-
  bind_rows(Encelia_elevation,
            Rafinesquia_elevation,
            Phacelia_elevation) %>% 
  st_drop_geometry() %>% 
  group_by(species, year) %>% 
  summarize(
    mean_e = mean(elevation),
    sd_e = sd(elevation),
    sample_size = n(),
    sem = sd_e / sqrt(sample_size),
    ci_upper_limit = mean_e + 1.96 * sem,
    ci_lower_limit = mean_e - 1.96 * sem,
  )
```

```{r}
elevation_point %>% 
  filter(!is.na(species)) %>% 
  ggplot() +
  geom_jitter(mapping = aes(x = year, y = elevation, color = year),
              alpha = .5) +
  geom_crossbar(
    data = all_elevation,
    mapping = aes(
      x = year,
      y = mean_e,
      ymax = ci_upper_limit,
      ymin = ci_lower_limit
    ),
    color = "black"
  ) +
  facet_wrap( ~ species) +
  guides(color = "none")
```

Figure 2. A jitter plot displaying the different elevations of the three plant species over the years 1995, 2005, and 2015.

```{r}
average_latitude <-
  all_data %>% 
  st_drop_geometry() %>% 
  group_by(species, year) %>% 
  summarize(
    mean_l = mean(latitude),
    sd_l = sd(latitude),
    sample_size = n(),
    sem = sd_l / sqrt(sample_size),
    ci_upper_limit = mean_l + 1.96 * sem,
    ci_lower_limit = mean_l - 1.96 * sem,
  ) 
```

```{r}
all_data %>% 
  ggplot() +
  geom_jitter(mapping = aes(x = year, y = latitude, color = year),
              alpha = .5) +
  geom_crossbar(
    data = average_latitude,
    mapping = aes(
      x = year,
      y = mean_l,
      ymax = ci_upper_limit,
      ymin = ci_lower_limit
    ),
    color = "black"
  ) +
  facet_wrap( ~ species) +
  guides(color = "none")
```

Figure 3. A jitter plot displaying the different latitudes of the three plant species over the years 1995, 2005, 2015.

## Discussion

## Literature Cited

1. Brian Maitner (2020). BIEN: Tools for Accessing the
   Botanical Information and Ecology Network Database. R package version 1.2.4.
   https://CRAN.R-project.org/package=BIEN
   
2. RStudio Team (2020). RStudio: Integrated Development Environment for
   R. RStudio, PBC, Boston, MA URL http://www.rstudio.com/
   
3. Hollister, J.W. (2020). elevatr: Access Elevation Data from Various APIs. 
   R package version 0.3.1.
   https://CRAN.R-project.org/package=elevatr/
   
4. R Core Team (2020). R: A language and environment for statistical computing. 
   R Foundation for Statistical Computing, Vienna, Austria. 
   URL https://www.R-project.org/





